---
Title: "Exploratory Data Analysis"
Author: "Jake Sauter"
output: 
  html_document: 
    toc: true
    toc_float: true
    df_print: kable
---


This script will cover the Exploratory Data Analysis (EDA) shown
in R for Data Science and may go a bit further

EDA is a cyclical process of 
1. Generating questions about data
2. Searching for answers in the data using summary stats
   and visualizations along with models
3. Generating more educated questions based on these
    insights to form even more refined insights 

```{r}
library(tidyverse)
library(magrittr)
```

Two questions that are often helpful to start with are 
1. What type of variation occurs within my variables
2. What type of covariation occurs between my variables

# Visualizing distributions 

What is the distribution of the cut of 
diamonds in my dataset

```{r}
diamonds %>% 
  count(cut)
```

hmm this is intriguing but maybe it would
be better to see this information in a graphic
form, such as a bargraph

```{r}
ggplot(diamonds) + 
  geom_bar(aes(x = cut))
```

Note that we used a bar graph and not a histogram here, 
bar charts are used for categorical values while histograms
involve the use of binning continuous variables

```{r}
ggplot(diamonds) + 
  geom_histogram(aes(x = carat), binwidth = 0.5)

smaller <- diamonds %>% 
  filter(carat < 3)

ggplot(smaller, aes(x = carat)) +
  geom_histogram(binwidth = 0.1)
```

If we want to overlay muliple distributions of variables
we can use frequency polygrams, a histogram though with a point
at the middle of each bar

```{r}
ggplot(smaller, aes(x = carat, color = cut)) + 
  geom_freqpoly(binwidth = 0.1)
```

Which values are the most common? Why?

Which values are rare? Why? Should we expect this?

Are there any unusual patterns? What might explain them?

```{r}
ggplot(smaller, aes(x = carat)) + 
  geom_histogram(binwidth = 0.01)
```

Why are there more diamonds at whole carasts and common
fractions of carats?

Why are there more diamonds slightly to the right of each peak
than there are slight to the left of each peak? 

What are there no diamonds bigger than 3 carats? 


Cluster of similar values suggest the existance of 
subgroups in the data, to understand these subgroups
ask the following questions: 

How are the observations within each cluster 
similar to eachother?

How are the observations in seperate clusters different 
from eachother?

How can the clusters be described?

Why might the appearance of the clusters be misleading?

```{r}
ggplot(faithful, aes(x = eruptions)) + 
  geom_histogram(binwidth = 0.25)
```

# Unusual Values

```{r}
ggplot(diamonds) + 
  geom_histogram(aes(x = y), binwidth = 0.5)
```

With this binwidth its hard for us to even see 
the outlier values near 0. To help us see these values 
we can zoom in using the coord_cartesian() ggplot2 
coordinate function

```{r}
ggplot(diamonds) + 
  geom_histogram(aes(x = y), binwidth = 0.5) + 
  coord_cartesian(ylim = c(0, 50))
```

Note that coord_cartesian() also has an xlim argument, 
and ggplot2 has seperate xlim() and ylim() functions 
that work slightly differently, by throwing away values outside
of the limits 

Zooming in on the y axis allows us to see three unusual values
~0, ~30, and ~60. Lets pluck them out with dplyr

```{r}
diamonds %>% 
  filter(y < 3 | y > 20) %>% 
  select(price, x, y, z) %>% 
  arrange(y)
```

The y variable measures one of the three dimensions of these
diamonds in mm. Notice that these values are 0 and could 
not be a possible measurement for these diamonds, incorrect
values. Also notice that these values are nearly 100 mm, very large
and their cost is not nearly proportional to the cost they 
should be for diamonds this size!

It is good practice to repeat analyses with and without 
the outliers. If they have minimal effect on the result and
we cannot figure out why they are there, it is reasonable
to replace them with missing values (probably removing them). 
However if they have a substantial effect on the results, we
should not just drop them without justification and need to 
figure out what caused them. (data entry error / etc ...)

If we do encounter these unusual values and don't think that 
they are valid, we have two options

1. Completely remove them 
2. Replace the unusual values with missing values 

```{r}
df <- diamonds %>% 
  filter(between(y, 3, 20))
```

This is most likely not the best option as although 
one measurement is not valid, it does not mean that 
all of the measurements for the observation are 
garbage 

It is best to replace these values with missing values, 
then to replace them with modified values (probably from 
interpolation) 

```{r}
df <- diamonds %>% 
  mutate(y = ifelse(y < 3 | y > 20, NA, y))
```

We can also use dplyr::case_when(), which is particularly 
useful when we want to create a new variable 

```{r}
x <- 1:50

case_when(
  x%% 35 == 0 ~ "fizz buzz", 
  x %% 5 == 0 ~ "fizz", 
  x %% 7 == 0 ~ "buzz", 
  TRUE ~ as.character(x)
)

diamonds %>% 
  mutate(color = case_when(
    color == "E" ~ "Eerlily blue", 
    color == "I" ~ "Indefinately black", 
    color == "J" ~ "Joyasley teal", 
    TRUE ~ "All diamonds are clear"
  )) %>% head(10)
```

Sometimes missing values mean something though, like 
in the case of missing dep_delay in nycflight13 
means that the flights were cancelled

```{r}
nycflights13::flights %>% 
  mutate(
    cancelled = is.na(dep_time), 
    sched_hour = sched_dep_time %/% 100, 
    sched_min = sched_dep_time %% 100, 
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot(aes(sched_dep_time)) + 
    geom_freqpoly(aes(color = cancelled), binwidth = 1/4)
```

This isn't a great way to summarise this data because 
there are much more non-cancelled flights than 
cancelled flights, we willcover this later. 

# Covariation of variables

## A Categorical and Continuous Variable 

It is common to want to explore the distribution 
of a continuous vriable broken down by categorical variable.

```{r}
ggplot(diamonds, aes(x = price)) + 
  geom_freqpoly(aes(color = cut), binwidth = 500)
```

Its hard to see the difference in distribution because 
the overall counts vary so much

```{R}
ggplot(diamonds) + 
  geom_bar(aes(x = cut))
```

To help in this scenario, we can change the y-axis
to be density instead of count. Density is count 
but standardized so that the area under each frequency 
polygram is one 

```{R}
ggplot(diamonds, aes(x = price, y = ..density..)) + 
  geom_freqpoly(aes(color = cut), binwidth = 500)
```

what are things like ..count.., ..proportion.., and ..density.. called
and where can I see documentation for them with a list of 
all possible items like it?

these special variables are returned from stat functions
in ggplot, such as stat_bin (called implicilty by 
geom_histogram)

these special variables can be found on the ggplot2 
cheat sheet: 
https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf

This plot is a little hard to read so lets take a look 
at displaying the same data in a box plot

```{R}
ggplot(diamonds, aes(x = cut, y = price)) + 
  geom_boxplot()
```

 Here we see much less information about the distrobution of the data, 
 though we can more easily compare the present distributions. The boxplot
 about supports the counterintuitive finding that better quality diamonds
 are cheaper on average!
 
 Notice that `cut` is an ordered factor, fair being worse than good 
 and so on. 
 
 Sometime we may want to reorder factor variables to behave a particular
 way. This can be achieved with the `reorder()` function.
 
```{r}
ggplot(mpg, aes(x = class, y = hwy)) + 
  geom_boxplot()
```
 

To make the trend more evident, lets use `reorder()` here. 

```{r}
ggplot(mpg) + 
  geom_boxplot(aes(x = reorder(class, hwy, FUN = median), y = hwy))
```


This plot might be better seen on its side for labelling reasons. 

```{r}
ggplot(mpg) + 
  geom_boxplot(aes(x = reorder(class, hwy, FUN = median), y = hwy)) + 
  coord_flip()
```


 **7.5.1.1 Question** What variable in the dataset is most important
 for predicting the price of a diamond? How is that variable correlated
 with cut? Why does the combination of these two relationships lead to
 lower quality diamonds being more expensive?
 
```{r}
vars <-
  diamonds %>% 
  lapply(is.numeric) %>% 
  unlist %>% which %>% names

vars %<>% syms

diamonds %>% 
  select(!!!vars) %>% 
  map(~cor(.x, diamonds$price)) %>% 
  unlist %>% abs %>% sort
```
 
 Here we see that `carat` is the most important predictive of price. Now how
 is this variable (carat) correlated with cut?
 
```{r}
diamonds %>% 
  group_by(cut) %>% 
  summarise(carat = median(carat))
```
 

So it seems that the lowest quality diamonds are more costly than the highest
quality diamonds as they tend to be much larger!


## Two Categorical Variables 

```{r}
ggplot(diamonds) + 
  geom_count(aes(x = cut, y = color))
```

Here we used the `geom_count` geom, though we can acheive the same
effect using `dplyr::count` along with `ggplot2::geom_tile`

```{r}
df <- diamonds %>% 
  count(color, cut) 

df %>% head(10)

df %>% 
  ggplot(aes(x = color, y = cut)) + 
  geom_tile(aes(fill = n))
```


I prefer the cleaner aesthetic of `ggplot2:geom_count`. 

## Two Continous Variables

The most simple method, a scatter plot 

```{r}
ggplot(diamonds) + 
  geom_point(aes(x = carat, y = price))
```


As the size of our dataset grows, scatterplots become less useful as overplotting
begins to takeover and piles form. We can fix this with the `alpha=.1` parameter.

```{r}
ggplot(diamonds) + 
  geom_point(aes(x = carat, y = price), alpha = .01)
```


Transparency can get dicey as well. This is when binning might be appropriate. We 
have seen `geom_histogram()` and `geom_freqpoly()` to bin one dimension. Now
we will see `geom_bin2d()` and `geom_hex()` for two dimensional binning

```{r}
ggplot(smaller) + 
  geom_bin2d(aes(x = carat, y = price))
```


This seems to accomplish what adding alpha levels to scatterplots attempts to, though by adding another dimension of the color scale becomes much cleaner. 

```{r}
# install.packages('hexbin')
ggplot(smaller) + 
  geom_hex(aes(x = carat, y = price))
```

`geom_hex` is slightly nicer and more R community themed.

Another option to tackle data of this form is to bin one of the continous
variable so that it acts like a categorical, then apply the previous methods
for a singular continous and categorical variable that we have learned about. 

```{r}
ggplot(smaller, aes(x = carat, y = price)) + 
  geom_boxplot(aes(group = cut_width(carat, 0.1)))
```


The `cut_width()` function used above divides `x` into bins of the provided
width. 

```{r}
cut_width(smaller$carat, 0.1) %>% first(5)
```

One final improvement on this 

```{r}
ggplot(smaller, aes(x = carat, y = price)) + 
  geom_boxplot(aes(group = cut_number(carat, 20)))

```


## Patterns and models 

If we ever spot a pattern, we should first ask ourselves a few questions.

* Could this pattern be due to random chance?
* How could we describe the relationship implied by the pattern?
* How strong is the relationship implied by the pattern?
* What other variables might affect this relationship?
* Does the relations ship if we look at individual subgroups of the data?



```{r}
ggplot(faithful) + 
  geom_point(aes(x = eruptions, y = waiting))
```


Patterns are one of the most useful tools of data scientist as they reveal 
covariation. If we think of variation as a a phenomenon that creates
uncertainty, covariation is a phenomenon that reduces it. 

If two variables covary, and this is due to causual relationship, then we can 
use one vairbale to predict the other. 


We will use the example of the `diamonds` dataset here. Its hard to understand
the relationship between cut and price because cut and carat, and carat and 
price are so tightly related. We will attempt to use a model to remove the 
very strong relationship between price and carat so we can explore the 
subtleties that remain 

```{r}
library(modelr)

mod <- lm(log(price) ~ log(carat), data = diamonds)

df <- diamonds %>% 
  add_residuals(mod) %>% 
  mutate(resid = exp(resid))

ggplot(df) + 
  geom_point(aes(x = carat, y = resid))
```


Once we remove the strong relationship between carat and price, we can see 
what we would expect in the relationship bettwen cut and price: relative
to their size, better quality diamonds are more expensive. 

```{r}
ggplot(df) + 
  geom_boxplot(aes(x = cut, y = resid))
```















