---
title: "Relational Data"
author: "Jake Sauter"
output: 
  html_document:  
    toc: true
    toc_float: true
    df_print: kable 
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  cache = FALSE, 
  warning = FALSE, 
  message = FALSE,
  hold = TRUE)

library(knitr)
library(kableExtra)

kable <- function(data, ...) {
  knitr::kable(data, digits=3, ...) %>% 
    kable_styling(bootstrap_options = c('striped', 
                                        'hover', 
                                        'responsive'),
                  full_width = FALSE, 
                  position = 'center', )
}

knit_print.data.frame <- function(x, ...) {
  res <- paste(c('', '', kable(head(x))), collapse = '\n')
  asis_output(res)
}

registerS3method('knit_print', 'data.frame', knit_print.data.frame)
```

The study of **relational data** is very important in the field of 
Data Science as very often complimenting data is stored in different
tables or different datasets. 

For this document we will use the `nycflights13` dataset to 
showcase the power of **R** and **SQL**

```{r libraries}
library(tidyverse)
library(nycflights13)
```

# The Data

`airlines` shows us the full carrier name from its abbreviated code. 

```{r}
airlines 
```

`airports` gives us information about each airport, identified by the 
`faa` airport code. 

```{r}
airports 
```

`planes` gives us information about each plane, identified by its `tailnum`

```{r}
planes 
```

`weather` gives the wather at each NYC airport for each hour. 

```{r}
weather 
```

These datasets are connected in many ways that we will see throughout 
this document.

# Keys

A **primary** key uniquely identifies an observation in its own table. 
An example of a primary key in the `planes` table woubl be the `tailnum`. 

A **foreign** key uniquely identifies an observation in another table, 
such as how `flights$tailnum` connects the `planes` table to the 
`flights` table (we will look more at this soon). 

A variable can both be a primary key **and** a foreign key. This can 
be seen in the current dataset as `origin` is part of the weather 
primary key, though also forms a relation between the `weather` table
and the `airport` table. 

Once we have identified primary keys in our tables, its good practice to 
verify that they do indeed **uniquely** identify each observation. One 
way to do this is to `count()` the primary keys.

```{r}
planes %>% 
  count(tailnum) %>% 
  filter(n > 1)

weather %>% 
  count(year, month, day, hour, origin) %>% 
  filter(n > 1)
```


Sometimes a table does not have an explicit primary key: each row 
is an observation but no combination of variables reliably identifies
it. For example we would think that the `flight` or `tailnum` variable is the primary key of the `flights` table, though there are mulitple flights for each
tailnumber a day.

```{r}
flights %>% 
  count(year, month, day, flight) %>% 
  filter(n > 1)
```


```{r}
flights %>% 
  count(year, month, day, tailnum) %>% 
  filter(n > 1)
```

In cases such as this, if there is no reliable primary key the 
best option might be to just make one. The key that we now
assign as our reliable key is called the **surrogate key**. 

# Mutating Joins

A **mutating join** allows us to combine variables from two tables. 
It first matches observations by their keys, then copies acroos variables
from one table to the other. 

```{r}
(flights2 <- flights %>% 
  select(year:day, hour, origin, dest, tailnum, carrier))
```

Imagine we want to add the full airline name to the `flights2` data, 
we can combine the `airlines` and `flights2` tables with a `left_join()`. 

```{r}
flights2 %>% 
  select(-origin, -dest) %>% 
  left_join(airlines, by = 'carrier')
```

This results in the addition of the `name` variable to the `flights2`
table. This is similar to using the following `mutate()` statement, 
thus why we call this kind of join a **mutating join**. 

```{r}
flights2 %>% 
  select(-origin, -dest) %>% 
  mutate(name = airlines$name[match(carrier, airlines$carrier)])
```

# Inner Join

```{r}
(x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     3, "x3"
))
(y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2",
     4, "y3"
))
```


The simplest type of join in the **inner join**. To be precise this 
should be called the inner **equijoin** because the keys are matched
by the **equ**ality operator. 

The output of an inner join is a table table that contains the key, 
the x values and the y values. 


```{r}
x %>% 
  inner_join(y, by = 'key')
```


# Outer joins

An inner join keeps oberservation that appear in both tables. An 
**outer join** keeps observations that appear in at least one 
table. There are three types of outer joins being 

* A **left join** keeps all observations in `x`
* A **right join** keeps all observations in `y`
* A **full join** keeps all observations in `x` and `y` 

If **duplicate keys** exist in one table this can be useful
such as in our airline name example, but if duplicate keys 
exist in both tables this is most likely an error and most 
querying engines will give **all possible combinations**, 
defined by the **Cartesian product**. 




























