---
title: "Strings with stringr"
author: "Jake Sauter"
output:
  html_document: 
    toc: true
    toc_float: true
    df_print: kable
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  cache = FALSE, 
  warning = FALSE, 
  message = FALSE,
  hold = TRUE)

library(knitr)
library(kableExtra)

kable <- function(data, ...) {
  knitr::kable(data, digits=3, ...) %>% 
    kable_styling(bootstrap_options = c('striped', 
                                        'hover', 
                                        'responsive'),
                  full_width = FALSE, 
                  position = 'center', )
}

knit_print.data.frame <- function(x, ...) {
  res <- paste(c('', '', kable(head(x))), collapse = '\n')
  asis_output(res)
}

registerS3method('knit_print', 'data.frame', knit_print.data.frame)
```


```{r libraries}
library(tidyverse)
library(stringr)
```

# Intro

```{r str-view-1}
x <- c("apple", "banana", "pear")

str_view(x, 'an')
```


`.` matches any character

```{r match-any}
str_view(x, '.a.')
```


To match `.`, we need to preface it with a `\`, but `\` is also a special character so we 
need to escape it as well!


```{r match-dot}
str_view('abc...', '\\.')
```


In order to match `\`, we first need to escape out escape, then need to do that again
to form the literal. 

```{r match-backslash}
str_view('a\\b', '\\\\')
```


# Anchors

* `^` matches the begining of a string. 
* `$` matches the end of a string. 

```{r}
x  <- c('apple', 'banana', 'pear')

str_view(x, '^a')

str_view(x, 'a$')
```


We can use both of these anchors to identify only complete strings. 

```{r}
x <- c('apple pie', 'apple', 'apple cake')

str_view(x, '^apple$')
```


# Classes and Alternatives

* `\d` matches any digit. 
* `\s` matches any whitespace (s for space). 
* `[abc]` matches a, b **or** c. 
* `[^abc]` matches anything **except** a, b or c.

We must remember when creating these regular expressions that we
have to escape the backslash before using it to form the regex expression. 
e.g. `\\d`. 

We can also use a character class containing a single character as an alternative
to a backslash escape for a single metacharacter in regex. 

```{r}
x <- c("abc", "a.c", "a*c", "a c")

str_view(x, 'a[.]c')

str_view(x, '.[*]c')

str_view(x, 'a[ ]')
```


This works for most but not all regex characters. The exceptions include `]`, `\`, `^` and `-`. 

We can use alternation to pick between one or more alternative patterns. 

```{r}
str_view(c('gray', 'grey'), 'gr(e|a)y')
```


# Repetition

* `?`: 0 or 1
* `+`: 1 or more 
* `*`: 0 or more

```{r}
x <- "1888 is the longest year in Roman numerals: MDCCCLXXXVIII"

str_view(x, "CC?")

str_view(x, 'C[LX]+')
```


* `{n}`: exactly n 
* `{n, }`: n or more
* `{,m}`: at most m
* `{n,m}` between n and m 

```{r}
str_view(x, 'C{2}')

str_view(x, 'C{2,}')

str_view(x, 'C{2,3}')
```


By default these matches are "greedy" and will match the longest possible string. We can change
this behavior to make matches "lazy" by putting a `?` after them. 

```{r}
str_view(x, 'C{2,3}?')

str_view(x, 'C[LX]+?')
```


# Grouping and backreferences

A numbered capture group. A capturing group stores the part of the string matches
by the part of the regular expression inside of the parentheses. We can refer to the 
same test as previously matched by the capturing group with backrefereces, like `\1` and `\2`. 

For exampe if we want to find all fruits with repeated pairs of letters.  

```{r}
str_view(fruit, '(..)\\1', match = TRUE)
```




```{r}
str_view('abaca', '(.).\\1.\\1')
```


# Detect matches

```{r}
x <- c('apple', 'banana', 'pear')

str_detect(x, 'e')
```

```{r}
sum(str_detect(words, '^t'))

str_view(words, '[aeiou]$', match = TRUE)

mean(str_detect(words, '[aeiou]$'))
```

Sometimes we may want to use multipe statements if we are trying to 
match a complex statement 

```{r}
no_vowels_1 <- !str_detect(words, '[aeiou]')

no_vowels_2 <- str_detect(words, '^[^aeiou]+$')

identical(no_vowels_1, no_vowels_2)
```

A common use of `str_match()` is to logical subset an array. 

```{r}
words[str_detect(words, 'x$')]
```

This can also be done with the `stringr::str_subset()` function.

```{r}
str_subset(words, 'x$')
```

We can also use `str_detect()` in `dplyr::filter()` calls. 

```{r}
df <- tibble(
  word = words, 
  i = seq_along(word)
)

df %>% 
  filter(str_detect(word, 'x$')) 
```

A variation on `str_detect()` is `str_count()`, telling how many matches
there are in a single string. 

```{r}
x <- c('apple', 'banana', 'pear')

str_count(x, 'a')

mean(str_count(words, '[aeiou]'))
```

This shows us that most common words have on average 2 vowels. 

A common design patter is to use `str_count()` with mutate. 

```{r}
df %>% 
  mutate(
    vowels = str_count(word, '[aeiou]'), 
    consonants = str_count(word, '[^aeiou]')
  )
```

Notice an interesting feature of regex of non-overlapping patterns. 

```{r}
str_count('abababa', 'aba')

str_view_all('abababa', 'aba')
```

Also take note here of the `_all` suffix. Most functions in `stringr`
come in a singular version, and an `_all` version. 

# Extract matches

```{r}
head(sentences)

colors <-  c('red', 'orange', 'yellow', 'green', 
              'blue', 'purple')

color_match <- str_c(colors, collapse = '|')
color_match
```


```{r}
has_color <- str_subset(sentences, color_match)
matches <- str_extract(has_color, color_match)
```

Note that `str_extract()` only extracts the first match. To see all matches
we have to use `str_extract_all()`

```{r}
str_extract_all(has_color, color_match) %>% unlist
```


```{r}
more <- sentences[str_count(sentences, color_match) > 1]
str_view_all(more, color_match)
```

```{r}
str_extract_all(more, color_match)
```

If we use `simplify = TRUE`, then `str_extract_all()` will return 
a matrix. 

```{r}
str_extract_all(more, color_match, simplify = TRUE)
```


# Grouped Matches

```{r}
noun <- '(a|the) ([^ ]+)'

(has_noun <- sentences %>% 
  str_subset(noun) %>% 
  head(10))

has_noun %>% 
  str_extract(noun)
```

While `str_extract()` gives us the complete match, `str_match()` will
give use each individual component. 

```{r}
has_noun %>% 
  str_match(noun)
```


If our data is in a tible, it is often easier to use `tidyr::extract()`. 

```{r}
tibble(sentences = sentences) %>% 
  tidyr::extract(sentences, 
                 c('article', 'noun'), 
                 '(a|the) ([^ ]+)', 
                 remove = FALSE) %>% 
  head(5)
```

# Replacing matches

`str_replace()` and `str_replace_all()` allow us to replace matches
with new strings. 

```{r}
x <- c('apple', 'pear', 'banana')

str_replace(x, '[aeiou]', '-')

str_replace_all(x, '[aeiou]', '-')
```

We can make multiple replacements by passing a named vector. 

```{r}
x <- c('1 house', '2 cars', '3 people')

str_replace_all(x, c('1' = 'one', '2' = 'two', '3' = 'three'))
```

Instead of replacing with a fixed string we can use backreferences
to insert components of the match. 

```{r}
sentences %>% 
  str_replace('([^ ]+) ([^ ]+) ([^ ]+)', '\\1 \\3 \\2') %>% 
  head(5)
```

# Splitting 

We use `str_split()` to split a string up into pieces. 

```{r}
sentences %>% 
  head(5) %>% 
  str_split(' ')
```

Due to the fact that each component might contain a different number
of pieces, a list is returned. If we are working with a length-1 vector
the easiest thing to do is to just extract the first element of the list. 

```{r}
'a|b|c|d' %>% 
  str_split('\\|') %>% 
  .[[1]]
```

Otherwise like the other `stringr` functions that return a list
we can use the `simplify = TRUE` option. 

```{r}
sentences %>% 
  head(5) %>% 
  str_split(' ', simplify = TRUE)
```

We can also request a maximum number of peices. 

```{r}
fields <- c("Name: Jake", "Country: USA", "Age: 22")
fields %>% str_split(': ', n = 1, simplify = TRUE)
```

Instead of splitting up strings by patters we can also split up by 
character, line, sentence and word `boundary()`s

```{r}
x <- 'This is a sentence. This is another sentence'

str_view_all(x, boundary('word'))
```

```{r}
str_split(x, ' ')[[1]]

str_split(x, boundary('word'))[[1]]
```

Notice how punctuation was removed when using `boundary(word)`. 

# Find matches

`str_locate(_all)?()` give us the starting and ending positions of
each match. This is useful wehn none of the included functions do 
extactly what we want to do. We can ues `str_locate()` to locate
the match and use `str_sub()` to replace it with our intended edit. 

We can always use other arguments of `regex()` to control 
details about the match. 

```{r}
bananas <- c("banana", "Banana", "BANANA")

str_view(bananas, 'banana')

str_view(bananas, regex('banana', ignore_case = TRUE))
```

# Other types of patterns

* `mulitline = TRUE` allows `^` and `$` to match the start and end of each line
rather tha the start and end of the complete string. 

```{r}
x <- "Line 1\nLine 2\nLine 3"

str_extract_all(x, '^Line')[[1]]

str_extract_all(x, regex('^Line', multiline = TRUE))[[1]]
```

* `comments = TRUE` allows us to use commands and whit space
to make complex regular expressions more understandable

```{r}
phone <- regex("
  \\(?     # optional opening parens
  (\\d{3}) # area code
  [) -]?   # optional closing parens, space, or dash
  (\\d{3}) # another three numbers
  [ -]?    # optional space or dash
  (\\d{3}) # three more numbers
  ", comments = TRUE)

str_match("514-791-8141", phone)
```

* `dotall = TRUE` allows `.` to match everything, including `\n`

There are **three** other functions that we can use **instead** of `regex()`

* `fixed()`: matches exactly the specified sequence of byes. It ignores
all special regular expressions and operates at a very low level. This allows
us to avioud complex escaping and can be much faster than regular expressions. 

```{r}
microbenchmark::microbenchmark(
  fixed = str_detect(sentences, fixed('the')), 
  regex = str_detect(sentences, 'the'), 
  times = 200)
```

`fixed()` can be up to **3x** faster than `regex()`! Though `fixed()`
may not work with non-English data.

* `coll()`: compare strings using standar **collocation** rules. This is
useful for doing case insensitive matching. Note that `coll()` takes a 
`locale` parameter that controls the rules that are used for comparing
characters as different parts of the world use different rules.

```{r}
i <- c("I", "İ", "i", "ı")

i

str_subset(i, coll("i", ignore_case = TRUE))

str_subset(i, coll("i", ignore_case = TRUE, locale = "tr"))
```

Both `fixed()` and `regex()` have ignore_case arguments, but they do not allow you to pick the locale: they always use the default `locale`. You can see what that is with the following code; more on stringi later.


```{r}
stringi::stri_locale_info()
```

`coll()` is relatively slow compared to `regex()` and `fixed()`. 

*`boundary()`: shown previously though we can use it with more functions


```{r}
x <- 'This is a sentence'

str_view_all(x, boundary('word'))
```
# Other uses of regex

* `apropos()` seaches all objects from the global environment. 

```{r}
apropos('replace')
```

* `dir()` lists all files in a directory. 

```{r}
head(dir(pattern = '\\.Rmd$'))
```

# Stringi 

`stringr` is built **ontop of** the `stringi` package. `stringi` is much
more comprehensive than `stringr` though has a steeper learning curve.
stringi has 234 functions to stringr’s 46. So if you can't find a function
to do what you need and think that it should be already implemented, 
check in `stringi`!
















