---
title: "Data Wrangling"
Author: "Jake Sauter"
output:
  html_document: 
    toc: true
    toc_float: true
    df_print: kable
---

# Tibbles

Tibbles *are* a `data.frame()`s, though have update functionality. They 
handily *do not convert strings to factors by default*. They also have
some better printing characteristics and easier formation practices.

```{r libraries}
library(tidyverse)
```

```{r as_tibble}
as_tibble(iris)
```

```{r tibble_creation}
tibble(
  x = 1:5, 
  y = 1, 
  z = x^2 + y
)
```

Its possible for a tibble to have column names that are not valid R variable
names (non-syntatic names). These names are surrounded in backticks

```{r special-col-names}
tb <- tibble(
  `:)` = 'smile', 
  ` `  = 'space', 
  `2000` = 'number'
)
```


Another way to create a tibble is with `tribble()`, short for 
**tr**anspose tibble. This function is is customised for data entry in
code.

```{r tribble}
tribble(
  ~x, ~y, ~z, 
  'a', 2, 3.6, 
  'b', 1, 8.5
)
```


The two main differences in `tibble` and `data.frame` are printing and subsetting.

## Printing

Tibbles have a refined print method that shows only the first 10 rows by default, 
along with all of the columns that fit on screen. 

```{r tibble-types}
tibble(
  a = lubridate::now() + runif(1e3) * 86400, 
  b = lubridate::today() + runif(1e3) * 30, 
  c = 1:1e3,
  d = runif(1e3),
  e = sample(letters, 1e3, replace = TRUE)
)
```


There are a few options that can allow more than the default 10 rows to be 
printed.

```{r inf-width-tibble-print}
nycflights13::flights %>% 
  print(n = 10, width = Inf)
```


The default print behavior can also be modified in the following way


```{r tibble-options}
options(tibble.print_max = 100, tibble.print_min = 3)
```

If a tibble has more than `100` rows, only `3` rows will be printed. 

```{r inf-tibble-print-option}
options(tibble.width = Inf)
```

Will always show all columns of tibbles.

A complete list of options can be found in `package?tibble`


We can also use RStudio's built-in data viewer to view tibbles.

```{r}
nycflights13::flights %>% 
  view
```


## Subsetting

We can use `$` and `[[` to extract data from a `data.frame`.

```{r}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

# Extract by name
df$x
df[["x"]]


# Extract by position
df[[1]]
```


To use these extraction methods in a pipe, we can use the special `.` operator.

```{r}
df %>% .$x

df %>% .[['x']]
```


Frustratingly some older code will not work with tibbles, and thus will need to use 
`as.data.frame()` in order to convert our tibble back into a data.frame.

```{r}
class(df) %>% print
class(as.data.frame(df)) %>% print
```


# Data Import

`readr::read_csv()` will work similarly to `base::read.csv()` though will return
a tibble


```{r}
library(microbenchmark)

csv_file <- '~/Documents/R/give_me_credit/data/cs-training.csv'

microbenchmark::microbenchmark(read.csv(csv_file), readr::read_csv(csv_file), times = 10)
```

This benchmark shows that `readr::read_csv()` takes *under half the time* as the `base::read.csv()` function, as 
well as read_csv() not casting strings to factors and a tibble with previously described niceties returned.


We can also specify inline csv to help in generating reproducible examples in R. 

```{r}
read_csv("a,b,c 
          1,2,3 
          4,5,6") %>% print

tribble(~a, ~b, ~c, 
         1,  2,  3, 
         4,  5,  6) %>% print
```


Sometime we want to skip the first few lines of a file 

```{r}
read_csv("first line of metdata
          second line of metadata
          x, y, z
          1, 2, 3", skip = 2)
```


Or we want to skip comments in a file that we have created.

```{r}
read_csv("#skip me, I'm a stupid comment, 
          x, y, z
          1, 2, 3", comment = "#")
```

If column names are not present in the file we can avoid using the first column as 
column names in the following way. 

```{r}
read_csv('1,2,3 \n 4,5,6', col_names = FALSE)
```

Or we can provide our own column names to use (thus also setting `col_names = FALSE` by doing so. 

```{r}
read_csv('1,2,3\n4,5,6', col_names = c('x', 'y', 'z'))
```

We can also tweak the `na` argument, specifying sentinel values in the file. 

```{r}
read_csv('a,b,c\n1,2,.', na = '.')
```

Some reasons to use `readr` over base R. 

* `readr` tends to be much quicker (~10x)
* `readr` produces tibbles, doesn't convert strings to factors, use row names, 
or change the column names if we want to. 
* `readr` helps produce reproducible code as base R functions inherit some behaviour
from the user's operating system.

*Note*: If we are after raw speed `data.frame::fread()` can be quite fast though does 
not fit into the tidyverse nearly as well as `readr` functions do 

## Parsing a vector

```{r}
parse_logical(c('TRUE', 'FALSE', 'NA')) %>% str 

parse_integer(c('1', '2', '3')) %>% str

parse_date(c('2010-01-01', '1997-04-14')) %>% str
```

These parsers can be helpful on their own but are also the building blocks of `readr`. 

Most functions in `readr` can take the arguments that we discussed earlier for `readr::read_csv()`


```{r}
parse_integer(c('1', '2', '3', '.'), na = '.') %>% str
```

If there is a parsing failure we will get a warning and all failures will be `NA`

```{r}
(x <- parse_integer(c('123', '345', 'abc', '123.45')))
```

If there are many parsing failures we will have to use the `problems()` function 
to see all of the errors. 

```{r}
problems(x)
```


Useful for writing numbers in a natural way

```{r}
parse_number('$100') %>% str

parse_number('20%') %>% str

parse_number('It cost $123.45') %>% str

parse_number('$123,432,323') %>% str
```

### Strings

Strings can also be parsed in different ways that might prove useful. 

```{r}
charToRaw('Hadley')
```

This is the *ASCII* code or the *American* Standard Code for Information Interchange.



```{r}
x1 <- "El Ni\xf1o was particularly bad this year"
x2 <- "\x82\xb1\x82\xf1\x82\xc9\x82\xbf\x82\xcd"

parse_character(x1, locale = locale(encoding = "Latin1")) %>% str
parse_character(x2, locale = locale(encoding = "Shift-JIS")) %>% str
```


So it is nice if we know the `locale` (aka the encoding scheme), but there
might be times where we don't know this encoding scheme. Fortunately we can use the handy function
`guess_encoding`

```{r}
guess_encoding(charToRaw(x1))
```


### Factors

```{r}
fruits <- c('apple', 'banana')

parse_factor(c('apple', 'banana', 'bananana'), levels = fruit)
```


### Dates, date-times and times

```{r}
parse_datetime('2010-10-01T2010') %>% str

parse_datetime('201001010') %>% str
```

```{r}
parse_date('2010-10-01') %>% str
```

```{r}
parse_time('01:10 am') %>% str

parse_time('20:10:01') %>% str
```


We can also specify our own format if we want with the `%` format.


```{r}
parse_date("01/02/15", "%m/%d/%y") %>% str

parse_date("01/02/15", "%d/%m/%y") %>% str

parse_date("01/02/15", "%y/%m/%d") %>% str
```


## Writing to a file
write_csv(df, 'df.csv')
df2 <- read_csv('df.csv')

df %>% print
df2 %>% print
```

Though the type of information that you are writing is lost when you 
write out to a csv.

```{r}
saveRDS(df, 'df.rds')
df2 <- readRDS('df.rds')

df %>% print
df2 %>% print
```


So the *.rds* file type will save the R object in original form, though the `feather` package
implements a fast binary file format that can be shared across programming languages.

```{r}
library(feather)

write_feather(df, 'df.feather')

df2 <- read_feather('df.feather')

df %>% print
df2 %>% print
```


Note that `haven` can read **SPSS** and **SAS** files. `readxl` reads **excel** files and `DBI` reads **sqlite** files.













```{r}
df <- df %>% 
  mutate(x = as.integer(x))


```



















